---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(width = 300)
```

# alloscore2

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/reichlab/alloscore2/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/reichlab/alloscore2/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`alloscore2` scores probabilistic forecasts by what they would have done with a
scarce resource.

Given forecasts for several targets and a shared budget `K`, it allocates that
budget to minimize expected loss, then scores the forecasts by the loss the
allocation actually incurred, relative to the loss of an allocation made by an
oracle that knew the outcomes in advance. A forecast that would have directed
resources well scores near zero; one that would have misdirected them scores
badly, however good its marginal calibration looks.

It reads [hubverse](https://hubverse.io) model output and target data directly.
It is a rewrite of the original
[alloscore](https://github.com/aaronger/alloscore) package by Aaron Gerding, and
reproduces its numerical results; see
`tests/testthat/test-legacy-equivalence.R`.

## Installation

```r
remotes::install_github("reichlab/alloscore2")
```

## Getting started

`vignette("alloscore2")` — [read it
online](https://reichlab.io/alloscore2/articles/alloscore2.html) — is the guided
tour: choosing which targets share a budget, reading the scores and their
per-target decomposition, varying the unit costs `w` and the asymmetry `alpha`,
and looking inside the optimization with the plotting functions. The sections
below are the short version.

## Scoring hubverse model output

`alloscore_model_out()` takes a `model_out_tbl` of quantile forecasts and the
matching `oracle_output`. The one thing you have to tell it is which task ID
columns enumerate the targets that *share* a budget — here the locations, so
each model, reference date and horizon is a separate allocation problem.

```{r example-score, message = FALSE}
library(alloscore2)
library(dplyr)

model_out <- hubExamples::forecast_outputs |>
  filter(output_type == "quantile")

scores <- alloscore_model_out(
  model_out_tbl = model_out,
  oracle_output = hubExamples::forecast_oracle_output,
  K = c(500, 1000, 2000),
  target_cols = "location",
  by = c("model_id", "K")
)
scores
```

The score is the realized loss less the oracle's, so smaller is better. Because
the budget is shared across locations, a model is rewarded for getting the
*relative* burden across locations right, not just each location on its own.

Passing `summarize = FALSE` returns one row per allocation problem, with a
nested `xdf` column holding each target's allocation, outcome and loss:

```{r example-detail}
detail <- alloscore_model_out(
  model_out_tbl = model_out,
  oracle_output = hubExamples::forecast_oracle_output,
  K = 1000,
  target_cols = "location",
  summarize = FALSE
)
detail$xdf[[1]]
```

## Allocating without scoring

`allocate_model_out()` stops after the allocation, which is useful when the
question is what a forecast implies you should do rather than how good it was:

```{r example-allocate}
allocate_model_out(
  model_out_tbl = filter(model_out, reference_date == "2022-11-19", horizon == 0),
  K = 1000,
  target_cols = "location"
) |>
  select(model_id, K, xdf) |>
  mutate(allocations = purrr::map(xdf, ~ select(.x, target_names, x))) |>
  select(-xdf) |>
  tidyr::unnest(allocations)
```

## Working outside the hubverse

The allocation machinery does not require hub data. Forecasts can be given as
any parametric family, with `add_pdqr_funs()` building the cdfs and quantile
functions from parameter columns:

```{r example-core}
forecasts <- add_pdqr_funs(
  tibble::tibble(
    target_names = c("a", "b", "c"),
    dist = "norm",
    mean = c(5, 8, 12),
    sd = c(1, 2, 3)
  ),
  types = c("p", "q")
)

alloscore(forecasts, y = c(4, 9, 11), K = c(10, 20, 30)) |>
  select(K, score, score_raw, score_oracle)
```

The loss is a generalized piecewise linear (pinball) loss, so `alpha` sets how
much worse it is to under-allocate than to over-allocate, `kappa` scales it, and
`w` gives the cost per unit allocated to each target. The
`stdize_news_params()`, `stdize_ou_params()` and `stdize_met_params()` helpers
convert the newsvendor, over/under-cost and cost-loss parameterizations into
`kappa` and `alpha`.

## Validation

The `zxh_tab2` and `zxh_tab3` datasets carry the published optima of two
budget-constrained multiproduct newsboy problems from Zhang, Xu and Hua (2009),
which `allocate()` reproduces. See `vignette("zxh-newsvendor")`.

## Code of Conduct

Please note that this package is released with a [Contributor Code of
Conduct](.github/CODE_OF_CONDUCT.md). By contributing to this project, you agree
to abide by its terms.

## Contributing

Interested in contributing back to the project? Please see our
[contributing guidelines](.github/CONTRIBUTING.md).
