---
title: "Scoring hubverse forecasts by what they would allocate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scoring hubverse forecasts by what they would allocate}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%"
)
```

```{r setup, message = FALSE}
library(alloscore2)
library(dplyr)
```

## Acknowledgment

This vignette adapts content from [the original `alloscore` package](https://github.com/aaronger/alloscore) written by Aaron Gerding. It was drafted by Claude Code (Opus 5) and reviewed for accuracy by Nick Reich.


## The idea

Most forecast scores judge each prediction on its own. An allocation score
instead asks what a whole vector of forecasts would have led you to *do* with a
scarce resource, and then charges you for the consequences.

Concretely: you have forecasts for $N$ targets and a budget $K$. You allocate
$x_i$ to target $i$, paying $w_i$ per unit, so that $\sum_i w_i x_i \le K$. When
the outcomes $y_i$ arrive, you incur a generalized piecewise linear loss

$$L_i(x_i, y_i) = \kappa_i \left[ (1-\alpha_i)(x_i - y_i)_+ + \alpha_i (y_i - x_i)_+ \right],$$

which with $\alpha_i$ near 1 punishes falling short much harder than
overshooting. 
Note also that $\kappa_i$ scales that loss — the cost per unit of error at target $i$ — so the pair splits into an over-cost $\kappa_i(1-\alpha_i)$ and an under-cost $\kappa_i\alpha_i$. 
Both $\alpha_i$ and $\kappa_i$ default to 1 in this package's computations.

The score is the loss your allocation incurred, minus the loss of
an allocation made by an oracle who knew the outcomes:

$$S = \sum_i L_i(x_i, y_i) - \sum_i L_i(x_i^{\text{oracle}}, y_i).$$

Because the budget is shared, this rewards getting the *relative* burden across
targets right.

## Looking at some forecast data

The forecast data in `hubExamples` is a standard hubverse `model_out_tbl`. We
filter to the quantile output type, which is the one `alloscore2` supports.

```{r data}
model_out <- hubExamples::forecast_outputs |>
  filter(output_type == "quantile")
oracle_output <- hubExamples::forecast_oracle_output

model_out |>
  select(model_id, reference_date, horizon, location, output_type_id, value) |>
  head()
```

### What's in the example data

`hubExamples::forecast_outputs` is a small slice of a FluSight-style hub: three
models forecasting weekly incident influenza hospitalizations in two states,
FIPS 25 (Massachusetts) and 48 (Texas), from two reference dates in the 2022–23
season, at horizons 0 to 3 weeks ahead. Each forecast has seven predictive quantiles.

```{r data-summary}
model_out |>
  select(model_id, target, location, reference_date, horizon, output_type_id) |>
  tidyr::pivot_longer(
    everything(),
    names_to = "column",
    values_transform = as.character
  ) |>
  summarise(
    n = n_distinct(value),
    values = paste(sort(unique(value)), collapse = ", "),
    .by = column
  )
```

That is $3 \times 1 \times 2 \times 2 \times 4 \times 7 = 336$ rows. Each
(model, reference date, horizon) combination therefore covers exactly two
locations, and those two are the set that will share a budget below.

The observed outcomes come from `hubExamples::forecast_oracle_output`, the full
FluSight target series for every location and week; `as_alloscore_df()` joins in
only the rows it needs, matching on `target`, `target_end_date` and `location`.
For the weeks these forecasts cover:

```{r observed}
oracle_output |>
  filter(
    output_type == "quantile",
    location %in% unique(model_out$location),
    target_end_date %in% unique(model_out$target_end_date)
  ) |>
  select(location, target_end_date, oracle_value) |>
  tidyr::pivot_wider(
    names_from = location, values_from = oracle_value, names_prefix = "loc_"
  )
```

Texas runs roughly three to fifteen times Massachusetts week to week, and the
two together range from about 1,300 to 2,600 hospitalizations. That is the scale
to keep in mind when reading budgets: $K = 500$ is severely binding, while
$K = 4000$ covers any week outright.


### Choosing what shares a budget

This is the one modeling decision the package cannot make for you. An allocation
problem is a *set* of targets competing for one budget, so you have to say which
task ID columns enumerate that set. Here we pool across locations, so
`target_cols = "location"`.

Everything else — `model_id` plus the remaining task IDs — becomes the
*allocation unit*, and one allocation problem is solved per combination. This follows the
same logic as a hubverse `compound_taskid_set`: the columns you name vary within
a group, and the rest are held constant. In the example below, we are preparing 
to make allocations based on the forecasts from the model in `model_id`, made 
for all the locations at a specific `reference_date`, `target`, `horizon` and 
`target_end_date`.

```{r as-alloscore-df}
adf <- as_alloscore_df(model_out, oracle_output, target_cols = "location")
adf
```

Each row in the table above holds one allocation problem. The `forecasts` column has a row per
target, with the predictive quantiles (`ps`, `qs`), a cdf `F` and quantile
function `Q` interpolated through them by
[distfromq](https://reichlab.io/distfromq/), and the observed outcome
`y`:

```{r one-unit}
adf$forecasts[[1]] |> select(location, y, ps, qs)
adf$forecasts[[1]]$F[[1]](100)
```

Above, `adf$forecasts[[1]]$F[[1]]` is the cumulative distribution function of the first target's predictive distribution, fit by `distfromq` through that row's ps/qs pairs — you call it at a value on the outcome scale and it returns the forecast probability that the outcome falls at or below that value.
So the printed value above is `F[[1]]` evaluated at the value of 100, and can be read as "what probability did this model put on location 25 seeing 100 or fewer?". And Q[[1]] is its inverse, taking a probability in $[0,1]$ back to the outcome scale.

Note that you rarely need to call `as_alloscore_df()` yourself — it is what
`alloscore_model_out()` does internally — but it is the place to look when a
join or a target set is not what you expected.

## Allocating without scoring

If you have a forecast and a budget but don't have the eventual observations to score against yet, you can use `allocate_model_out()`, which stops after the allocation step. This is useful when the
question is what a forecast implies you should do rather than how good it was:

```{r example-allocate}
allocate_model_out(
  model_out_tbl = filter(model_out, reference_date == "2022-11-19", horizon == 0),
  K = 1000,
  target_cols = "location"
) |>
  select(model_id, K, xdf) |>
  mutate(allocations = purrr::map(xdf, ~ select(.x, target_names, x))) |>
  select(-xdf) |>
  tidyr::unnest(allocations)
```

In the above output, `x` is the column giving you the suggested allocations from each model for each location.


## Allocation scores

However, the focus of this package is on scoring different choices of allocations based on a model's forecast. Here is code to do that, for a variety of different budgets $K$.

```{r score}
scores <- alloscore_model_out(
  model_out_tbl = model_out,
  oracle_output = oracle_output,
  K = c(500, 1000, 2000, 4000),
  target_cols = "location",
  by = c("model_id", "K")
)
scores
```

Smaller scores are better. Note the `ytot` column: the total outcome across the pooled
locations. Once $K$ exceeds it the oracle can cover everything and its loss
falls to zero, so differences between models become differences in how well they
spent a budget that was never really binding.

Note that scores are differences of losses against an oracle, so they in theory should never be below zero. However, small negative values can occur in practice (see results above) as an artifact of the `eps_K` argument to `allocate()` or `alloscore_model_out()`, which lets the forecaster's allocation overspend the budget by up to 1% while the oracle is held to $K$ exactly, and it shrinks toward zero as you tighten that tolerance.

### Digging into per-target details

With `summarize = FALSE` you get one row per allocation problem and budget, plus
a nested `xdf` holding the allocation, outcome and loss for each target.

```{r detail}
detail <- alloscore_model_out(
  model_out_tbl = model_out,
  oracle_output = oracle_output,
  K = 2000,
  target_cols = "location",
  summarize = FALSE
)
detail |> select(model_id, reference_date, horizon, K, score, xdf)
detail$xdf[[2]]
```

The `components` column is each target's contribution to the overall allocation score. 
In the above case, the model's total shortfall was 163 against the oracle's 150, and the 13-unit gap (the `score` in row 2 of the detail tibble) is precisely what it overspent on location 25 (234 against a need of 221). 
With alpha = 1, a binding budget makes the score simply the budget wasted on targets that didn't need it — overshooting 25 is not penalized directly (its `components_raw` is 0), but it is penalized because those units were unavailable to location 48.

## Costs that differ by target

`w` is the cost of a unit allocated to a target. It can be a scalar, a vector
named by target, or the name of a column of the model output. Making one
location more expensive shifts the allocation away from it:

```{r weights}
one_round <- model_out |>
  filter(reference_date == "2022-11-19", horizon == 0)

bind_rows(
  allocate_model_out(one_round, K = 1000, target_cols = "location") |>
    mutate(weighting = "equal cost"),
  allocate_model_out(
    one_round,
    K = 1000, target_cols = "location", w = c("25" = 1, "48" = 5)
  ) |>
    mutate(weighting = "location 48 five times dearer")
) |>
  select(weighting, model_id, xdf) |>
  mutate(alloc = purrr::map(xdf, ~ select(.x, target_names, x))) |>
  select(-xdf) |>
  tidyr::unnest(alloc) |>
  tidyr::pivot_wider(names_from = target_names, values_from = x)
```

In the results shown above, the total "cost" of the allocation shown in the rightmost rows are always the same (the sum of the value in column `25` plus five times the sum of the value in `48` is 1000 +/- 1%) but because it is more expensive to allocate to location 48 there are fewer resources deployed there in rows 4-6.

## How much it hurts to fall short

`alpha` is the quantile level of the underlying pinball loss, and it encodes how
asymmetric the consequences are. The unconstrained optimum for a target is its
`alpha`-quantile, so raising `alpha` buys more insurance against shortfalls.

```{r alpha}
purrr::map(
  c(0.5, 0.8, 0.95, 1),
  function(a) {
    alloscore_model_out(
      one_round,
      oracle_output,
      K = 2000,
      target_cols = "location",
      alpha = a,
      by = "model_id"
    ) |>
      mutate(alpha = a)
  }
) |>
  purrr::list_rbind() |>
  select(alpha, model_id, score) |>
  tidyr::pivot_wider(names_from = model_id, values_from = score)
```

The above results show that PSI-DICE allocates best at every asymmetry level, Flusight-baseline worst — by a widening margin as alpha rises, since its tight, low predictive distributions leave shortfalls that grow costlier — while at alpha = 1 all three score zero because the unbounded optimum makes each spend the full budget, which at $K = 2000$ covers this week's total of 1309 outright.

## Looking inside the optimization

The allocation is found by bisecting on a single Lagrange multiplier
$\lambda$, the shadow price of the budget. At the optimum every target that
receives anything has the same marginal expected benefit, and each allocation is
that target's predictive quantile at a level discounted by $\lambda$.
`plot_iterations()` shows that search.

```{r iterations, fig.height = 5}
alloc <- allocate_model_out(one_round, K = 2000, target_cols = "location")
plot_iterations(dplyr::filter(alloc, model_id == "PSI-DICE"), K_to_plot = 2000)
```

The upper panel stacks the allocations, with the horizontal line at the budget;
the lower panel tracks $\lambda$ and the interval bracketing it.

Loosely, $\lambda$ is what one more unit of budget would be worth — how much expected loss it would save — so it is large when the budget is tight and falls to zero once the budget stops binding.
With equal costs and weights it also has a direct reading: instead of covering each target up to its $\alpha$-quantile, you can only afford its $(\alpha - \lambda)$-quantile, and $\lambda$ is the same for every target that gets anything. At $\alpha = 0.95$ and $\lambda = 0.1$, each location is stocked to its 85th percentile rather than its 95th.

## Score decomposition

`plot_components_slim()` shows which targets a score came from. Over a grid of
budgets it stacks each target's contribution.

```{r components}
scored <- alloscore_model_out(
  model_out_tbl = one_round,
  oracle_output = oracle_output,
  K = seq(250, 4000, by = 250),
  target_cols = "location",
  summarize = FALSE
)
plot_components_slim(
  scored,
  model_col_name = "model_id",
  show_oracle = FALSE
)
```

## Scoring outside the hubverse

The core functions take forecasts as a data frame with one row per target and
list columns of cdfs and quantile functions, which `add_pdqr_funs()` builds from
parameter columns for any distribution R knows.

```{r core}
forecasts <- add_pdqr_funs(
  tibble::tibble(
    target_names = c("a", "b", "c"),
    dist = "norm",
    mean = c(5, 8, 12),
    sd = c(1, 2, 3)
  ),
  types = c("p", "q")
)
alloscore(forecasts, y = c(4, 9, 11), K = c(10, 20, 30)) |>
  select(K, score, score_raw, score_oracle, ytot)
```

For Monte Carlo work — one allocation scored against many simulated outcome
vectors — `slim()` strips the allocation down to what scoring needs, and
`alloscore()` then accepts a list of outcome vectors.

```{r monte-carlo}
alloc <- slim(allocate(forecasts, K = c(10, 20, 30)))
ys <- withr::with_seed(42, {
  purrr::map(1:50, function(i) {
    rlang::set_names(rexp(3, rate = 1 / 8), c("a", "b", "c"))
  })
})
mc <- alloscore(alloc, ys)
mc |>
  select(samp, scores) |>
  tidyr::unnest(scores) |>
  group_by(K) |>
  summarise(mean_score = mean(score), sd_score = sd(score), .groups = "drop")
```

## Caveats

* Only the `quantile` output type is supported. Scoring `sample` output, which
  would let the allocation respect dependence across targets rather than working
  from marginals, is not implemented.
* The budget constraint is enforced only to within `eps_K`, 1% by default. An
  allocation may therefore overspend slightly, and since the oracle is held to
  the budget exactly, a score can come out marginally negative. Tighten `eps_K`
  if that matters for your application.
* `oracle_allocate()` spends the whole budget even when the oracle has no use
  for it, so the oracle allocation it reports is inflated for generous budgets.
  The oracle's *loss*, and hence the score, is unaffected.
