---
title: "Scoring hubverse forecasts by what they would allocate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scoring hubverse forecasts by what they would allocate}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%"
)
```

```{r setup, message = FALSE}
library(alloscore2)
library(dplyr)
```

## The idea

Most forecast scores judge each prediction on its own. An allocation score
instead asks what a whole vector of forecasts would have led you to *do* with a
scarce resource, and then charges you for the consequences.

Concretely: you have forecasts for $N$ targets and a budget $K$. You allocate
$x_i$ to target $i$, paying $w_i$ per unit, so that $\sum_i w_i x_i \le K$. When
the outcomes $y_i$ arrive, you incur a generalized piecewise linear loss

$$L_i(x_i, y_i) = \kappa_i \left[ (1-\alpha_i)(x_i - y_i)_+ + \alpha_i (y_i - x_i)_+ \right],$$

which with $\alpha_i$ near 1 punishes falling short much harder than
overshooting. The score is the loss your allocation incurred, minus the loss of
an allocation made by an oracle who knew the outcomes:

$$S = \sum_i L_i(x_i, y_i) - \sum_i L_i(x_i^{\text{oracle}}, y_i).$$

Because the budget is shared, this rewards getting the *relative* burden across
targets right. A model can be beautifully calibrated at every location on its
own and still allocate badly if it misjudges which locations matter most.

## Scoring a hub's model output

The forecast data in `hubExamples` is a standard hubverse `model_out_tbl`. We
filter to the quantile output type, which is the one `alloscore2` supports.

```{r data}
model_out <- hubExamples::forecast_outputs |>
  filter(output_type == "quantile")
oracle_output <- hubExamples::forecast_oracle_output

model_out |>
  select(model_id, reference_date, horizon, location, output_type_id, value) |>
  head()
```

### Choosing what shares a budget

This is the one modeling decision the package cannot make for you. An allocation
problem is a *set* of targets competing for one budget, so you have to say which
task ID columns enumerate that set. Here we pool across locations, so
`target_cols = "location"`.

Everything else — `model_id` plus the remaining task IDs — becomes the
*allocation unit*, and one problem is solved per combination. This follows the
same logic as a hubverse `compound_taskid_set`: the columns you name vary within
a group, and the rest are held constant.

```{r as-alloscore-df}
adf <- as_alloscore_df(model_out, oracle_output, target_cols = "location")
adf
```

Each row holds one allocation problem. The `forecasts` column has a row per
target, with the predictive quantiles (`ps`, `qs`), a cdf `F` and quantile
function `Q` interpolated through them by
[distfromq](https://github.com/reichlab/distfromq), and the observed outcome
`y`:

```{r one-unit}
adf$forecasts[[1]] |> select(location, y, ps, qs)
adf$forecasts[[1]]$F[[1]](100)
```

You rarely need to call `as_alloscore_df()` yourself — it is what
`alloscore_model_out()` does internally — but it is the place to look when a
join or a target set is not what you expected.

### Scores

```{r score}
scores <- alloscore_model_out(
  model_out_tbl = model_out,
  oracle_output = oracle_output,
  K = c(500, 1000, 2000, 4000),
  target_cols = "location",
  by = c("model_id", "K")
)
scores
```

Smaller is better. Note the `ytot` column: the total outcome across the pooled
locations. Once $K$ exceeds it the oracle can cover everything and its loss
falls to zero, so differences between models become differences in how well they
spent a budget that was never really binding.

### Per-target detail

With `summarize = FALSE` you get one row per allocation problem and budget, plus
a nested `xdf` holding the allocation, outcome and loss for each target.

```{r detail}
detail <- alloscore_model_out(
  model_out_tbl = model_out,
  oracle_output = oracle_output,
  K = 2000,
  target_cols = "location",
  summarize = FALSE
)
detail |> select(model_id, reference_date, horizon, K, score)
detail$xdf[[1]]
```

The `components` column is each target's contribution to the score. A positive
value means the forecast sent too little or too much to that target relative to
what the oracle would have done.

## Costs that differ by target

`w` is the cost of a unit allocated to a target. It can be a scalar, a vector
named by target, or the name of a column of the model output. Making one
location more expensive shifts the allocation away from it:

```{r weights}
one_round <- model_out |>
  filter(reference_date == "2022-11-19", horizon == 0)

bind_rows(
  allocate_model_out(one_round, K = 1000, target_cols = "location") |>
    mutate(weighting = "equal cost"),
  allocate_model_out(
    one_round,
    K = 1000, target_cols = "location", w = c("25" = 1, "48" = 5)
  ) |>
    mutate(weighting = "location 48 five times dearer")
) |>
  select(weighting, model_id, xdf) |>
  mutate(alloc = purrr::map(xdf, ~ select(.x, target_names, x))) |>
  select(-xdf) |>
  tidyr::unnest(alloc) |>
  tidyr::pivot_wider(names_from = target_names, values_from = x)
```

## How much it hurts to fall short

`alpha` is the quantile level of the underlying pinball loss, and it encodes how
asymmetric the consequences are. The unconstrained optimum for a target is its
`alpha`-quantile, so raising `alpha` buys more insurance against shortfalls.

```{r alpha}
purrr::map(
  c(0.5, 0.8, 0.95, 1),
  function(a) {
    alloscore_model_out(
      one_round,
      oracle_output,
      K = 2000,
      target_cols = "location",
      alpha = a,
      by = "model_id"
    ) |>
      mutate(alpha = a)
  }
) |>
  purrr::list_rbind() |>
  select(alpha, model_id, score) |>
  tidyr::pivot_wider(names_from = model_id, values_from = score)
```

## Looking inside the optimization

The allocation is found by bisecting on a single Lagrange multiplier
$\lambda$, the shadow price of the budget. At the optimum every target that
receives anything has the same marginal expected benefit, and each allocation is
that target's predictive quantile at a level discounted by $\lambda$.
`plot_iterations()` shows that search.

```{r iterations, fig.height = 5}
alloc <- allocate_model_out(one_round, K = 2000, target_cols = "location")
plot_iterations(dplyr::filter(alloc, model_id == "PSI-DICE"), K_to_plot = 2000)
```

The upper panel stacks the allocations, with the horizontal line at the budget;
the lower panel tracks $\lambda$ and the interval bracketing it.

## Score decomposition

`plot_components_slim()` shows which targets a score came from. Over a grid of
budgets it stacks each target's contribution.

```{r components}
scored <- alloscore_model_out(
  model_out_tbl = one_round,
  oracle_output = oracle_output,
  K = seq(250, 4000, by = 250),
  target_cols = "location",
  summarize = FALSE
)
plot_components_slim(
  scored,
  model_col_name = "model_id",
  show_oracle = FALSE
)
```

## Scoring outside the hubverse

The core functions take forecasts as a data frame with one row per target and
list columns of cdfs and quantile functions, which `add_pdqr_funs()` builds from
parameter columns for any distribution R knows.

```{r core}
forecasts <- add_pdqr_funs(
  tibble::tibble(
    target_names = c("a", "b", "c"),
    dist = "norm",
    mean = c(5, 8, 12),
    sd = c(1, 2, 3)
  ),
  types = c("p", "q")
)
alloscore(forecasts, y = c(4, 9, 11), K = c(10, 20, 30)) |>
  select(K, score, score_raw, score_oracle, ytot)
```

For Monte Carlo work — one allocation scored against many simulated outcome
vectors — `slim()` strips the allocation down to what scoring needs, and
`alloscore()` then accepts a list of outcome vectors.

```{r monte-carlo}
alloc <- slim(allocate(forecasts, K = c(10, 20, 30)))
ys <- withr::with_seed(42, {
  purrr::map(1:50, function(i) {
    rlang::set_names(rexp(3, rate = 1 / 8), c("a", "b", "c"))
  })
})
mc <- alloscore(alloc, ys)
mc |>
  select(samp, scores) |>
  tidyr::unnest(scores) |>
  group_by(K) |>
  summarise(mean_score = mean(score), sd_score = sd(score), .groups = "drop")
```

## Caveats

* Only the `quantile` output type is supported. Scoring `sample` output, which
  would let the allocation respect dependence across targets rather than working
  from marginals, is not implemented.
* The budget constraint is enforced only to within `eps_K`, 1% by default. An
  allocation may therefore overspend slightly, and since the oracle is held to
  the budget exactly, a score can come out marginally negative. Tighten `eps_K`
  if that matters for your application.
* `oracle_allocate()` spends the whole budget even when the oracle has no use
  for it, so the oracle allocation it reports is inflated for generous budgets.
  The oracle's *loss*, and hence the score, is unaffected.
