---
title: "Validation against published newsvendor optima"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validation against published newsvendor optima}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%"
)
```

```{r setup, message = FALSE}
library(alloscore2)
library(dplyr)
library(purrr)
```

## Acknowledgment

This vignette adapts content from [the original `alloscore` package](https://github.com/aaronger/alloscore) written by Aaron Gerding, specifically from [a vignette](https://github.com/aaronger/alloscore/blob/main/vignettes/zxh_demo.Rmd) that shows how to reproduce results from Zhang, Xu and Hua (2009).

This vignette was drafted by Claude Code (Opus 5) and reviewed for accuracy by Nick Reich.

## Why the newsvendor problem

The optimization inside [allocate()] is not specific to forecast evaluation. It
solves a budget-constrained separable convex program, and one classical instance
of that is the *multiproduct newsboy problem with a budget constraint*: choose
stock levels for several products, under uncertain demand, subject to a limit on
total purchasing cost.

That is useful here because the problem has a published literature with reported
optima. Zhang, Xu and Hua (2009) give two worked examples, and this package
ships their data in `zxh_tab2` and `zxh_tab3`. Reproducing their `Opt` column
checks `allocate()` against an external source rather than against the package's
own history.

```{r data}
zxh_tab2
```

## From newsvendor costs to gpl parameters

The newsvendor parameterization uses three costs per product: `c` per unit
purchased, `v` per unit of unmet demand, and `h` per unit left over. The gpl
loss instead uses a scale `kappa` and a level `alpha`.
`stdize_news_params()` converts between them, giving

$$\kappa = h + v, \qquad \alpha = \frac{v - c}{h + v}.$$

The `alpha` that falls out is the newsvendor's familiar critical ratio, and the
unconstrained optimum is the `alpha`-quantile of demand.

```{r reparam}
zxh_norm <- zxh_tab2 |>
  mutate(stdize_news_params(ax = c, a_minus = v, a_plus = h), .after = c) |>
  as_tibble()
zxh_norm |> select(Product, c, v, h, kappa, alpha, q_zxh)
```

## Building the demand distributions

Demand is Normal here, so `add_pdqr_funs()` builds the cdf and quantile function
from `mean` and `sd` columns. Evaluating each quantile function at its own
`alpha` should recover the critical quantile that Zhang, Xu and Hua print in
`q_zxh`.

```{r forecasts}
zxh_norm_forecasts <- zxh_norm |>
  rename(mean = mu, sd = sigma) |>
  add_pdqr_funs(dist = "norm", types = c("p", "q")) |>
  mutate(q_ours = map2_dbl(Q, alpha, function(Q, alpha) Q(alpha)))

zxh_norm_forecasts |>
  select(Product, alpha, q_zxh, q_ours) |>
  mutate(difference = q_ours - q_zxh)
```

## Allocating under the budget

Zhang, Xu and Hua use a budget of 2500, with the unit cost `c` as the weight in
the constraint. Passing `w = "c"` tells `allocate()` to take the weights from
that column, and `target_names = "Product"` names the targets from another.

```{r allocate}
allo_norm <- allocate(
  zxh_norm_forecasts,
  w = "c",
  K = 2500,
  target_names = "Product"
)

comparison <- zxh_norm_forecasts |>
  select(Product, c, GIM, Opt) |>
  left_join(allo_norm$xdf[[1]], by = "Product") |>
  select(Product, c, GIM, Opt, ours = x) |>
  mutate(difference = ours - Opt)
comparison
```

Two things to note. Only 6 of the 17 products are stocked at all, and they are
exactly the 6 that Zhang, Xu and Hua stock — the budget binds hard enough that
most products are shut out entirely, which is a useful test of the boundary
handling. And the agreement is to within about a twentieth of a unit:

```{r accuracy}
c(
  max_abs_difference = max(abs(comparison$difference)),
  budget_spent = sum(comparison$c * comparison$ours),
  budget_available = 2500,
  n_stocked_ours = sum(comparison$ours > 1e-8),
  n_stocked_zxh = sum(comparison$Opt > 0)
)
```

The budget is spent to slightly over 2500. That is the `eps_K` tolerance, 1% by
default: the bisection stops once the constraint is satisfied to within that
relative tolerance rather than exactly.

## Comparing objective values

The comparison that really matters is not the allocation vector but the value of
the objective, since the problem can have near-flat optima.
`exp_gpl_loss_fun()` builds the expected loss for one product, including the
purchasing cost as an offset.

```{r objective}
objective <- function(x) {
  sum(pmap_dbl(
    list(
      zxh_norm_forecasts$F,
      zxh_norm_forecasts$kappa,
      zxh_norm_forecasts$alpha,
      zxh_norm_forecasts$mean,
      zxh_norm_forecasts$c,
      x
    ),
    function(F, kappa, alpha, mean, c, x) {
      exp_gpl_loss_fun(F = F, kappa = kappa, alpha = alpha, offset = c * mean)(x)
    }
  ))
}

c(
  objective_at_zxh_opt = objective(comparison$Opt),
  objective_at_ours = objective(comparison$ours),
  objective_at_gim = objective(comparison$GIM)
)
```

Our objective value is marginally lower than at the published optimum, which is
what the slight overspend buys. Both are below the earlier
Abdel-Malek and Montanari solution in the `GIM` column.

## The search

`plot_iterations()` shows the bisection on the shadow price of the budget. With
the default tolerances it takes 15 steps, the last few of which barely move the
allocation.

```{r iterations, fig.height = 5}
plot_iterations(allo_norm, num_targets_to_color = 6)
```

## A bounded support

The second example uses Beta demand on $[x_{\min}, x_{\max}]$ rather than the
whole real line. `add_pdqr_funs()` handles that with `trans` and `trans_inv`,
which are composed onto the cdf and the quantile function respectively.

```{r beta}
zxh_beta_forecasts <- zxh_tab3 |>
  mutate(stdize_news_params(ax = c, a_minus = v, a_plus = h), .after = c) |>
  as_tibble() |>
  rename(shape1 = Balpha, shape2 = Bbeta) |>
  add_pdqr_funs(
    types = c("p", "q"),
    dist = "beta",
    trans = function(x, x_min, x_max) (x - x_min) / (x_max - x_min),
    trans_inv = function(q, x_min, x_max) x_min + (x_max - x_min) * q
  )

allo_beta <- allocate(
  zxh_beta_forecasts,
  w = "c",
  K = 6500,
  target_names = "Product"
)

zxh_beta_forecasts |>
  select(Product, c, x_min, x_max, GIM, Opt) |>
  left_join(allo_beta$xdf[[1]], by = "Product") |>
  select(Product, c, x_min, x_max, GIM, Opt, ours = x) |>
  mutate(difference = ours - Opt)
```

Every product is stocked here, and each allocation lands inside its own support.

```{r beta-accuracy}
beta_cmp <- zxh_beta_forecasts |>
  select(Product, c, Opt, x_min, x_max) |>
  left_join(allo_beta$xdf[[1]], by = "Product")
c(
  max_abs_difference = max(abs(beta_cmp$x - beta_cmp$Opt)),
  budget_spent = sum(beta_cmp$c * beta_cmp$x),
  budget_available = 6500,
  all_within_support = all(beta_cmp$x >= beta_cmp$x_min & beta_cmp$x <= beta_cmp$x_max)
)
```

## Other parameterizations

Two more helpers convert application parameterizations into `kappa` and `alpha`:
`stdize_ou_params()` takes explicit over- and under-prediction costs, and
`stdize_met_params()` takes the cost-loss ratio used in the weather
decision-analysis literature.

```{r other-params}
rbind(
  newsvendor = stdize_news_params(ax = 4, a_minus = 7, a_plus = 1),
  over_under = stdize_ou_params(O = 1, U = 3),
  cost_loss = stdize_met_params(C = 2, L = 10)
)
```

## References

B. Zhang, X. Xu, and Z. Hua. "A binary solution method for the multiproduct
newsboy problem with budget constraint." *International Journal of Production
Economics*, 117(1):136-141, 2009.

L. Abdel-Malek and R. Montanari. "An analysis of the multi-product newsboy
problem with a budget constraint." *International Journal of Production
Economics*, 97(3):296-307, 2005.
